---
title: 'Data preprocessing'
output:
  html_document:
    code_folding: 'hide'
---

```{r load_packages, echo=TRUE, include=FALSE}
library(tidyverse)
library(biomaRt)
library(knitr)
library(expss)
```

<br>

### Load SFARI file and add BioMart information

<br>

**Note:** I was originally using the ensembl archive from feb2014 to get the Ensembl IDs because that's the version used for Gandal's dataset and I was trying to get as many matches as possible, but I think this was returning me outdated information (gene_biotype) for some of the genes, so now, I'm going to:

1. Query with the current version of BioMart

2. Query with the feb2014 archive version to complement the Ensembl IDs that are no longer present in the new version (using only the current version, 9 genes cannot be matched to the Gandal dataset, when with the 2014 version only one was missing)

```{r load_and_annotate_data, message=FALSE, warning=FALSE}
# Load SFARI database
SFARI_genes = read_csv('./../Data/SFARI_genes_08-29-2019.csv')

# Add Ensembl IDs

# Query current version of BioMart
getinfo = c('hgnc_symbol','ensembl_gene_id','gene_biotype')
mart = useMart(biomart='ENSEMBL_MART_ENSEMBL', dataset='hsapiens_gene_ensembl', host='jan2020.archive.ensembl.org')
datGenes_now = getBM(attributes=getinfo, filters='hgnc_symbol', values=SFARI_genes$`gene-symbol`, mart=mart)

# Query feb2014  archive
mart = useMart(biomart='ENSEMBL_MART_ENSEMBL', dataset='hsapiens_gene_ensembl', host='feb2014.archive.ensembl.org')
datGenes_2014 = getBM(attributes=getinfo, filters='hgnc_symbol', values=SFARI_genes$`gene-symbol`, mart=mart)
datGenes_2014 = datGenes_2014 %>% filter(!ensembl_gene_id %in% datGenes_now$ensembl_gene_id)

# Merge results
datGenes = rbind(datGenes_now, datGenes_2014) %>% data.frame %>% arrange(hgnc_symbol)

# Merge SFARI_genes with Ensembl IDs
SFARI_genes = SFARI_genes %>% left_join(datGenes, by=c('gene-symbol'='hgnc_symbol')) %>% mutate(ID = ensembl_gene_id)

rm(getinfo, mart, datGenes_now, datGenes_2014)
```

<br>

### Improving BioMart results

<br>

1. The feb2014 archive version includes some IDs of the form 'LRG_n'
```{r LRG_IDs}
SFARI_genes$ensembl_gene_id[grepl('LRG_', SFARI_genes$ensembl_gene_id)]
```

These always have an alternative regular Ensembl ID, so we can just filter them out of the dataset without losing information
```{r}
SFARI_genes = SFARI_genes %>% filter(!grepl('LRG_',ensembl_gene_id))
```

<br>

2. Genes that didn't get a match (only 2 in common with the previous BioMart version)
```{r}
print(paste0(sum(is.na(SFARI_genes$ID)),' Genes didn\'t return any matches: ', 
             paste(sort(SFARI_genes$`gene-symbol`[is.na(SFARI_genes$ID)]), collapse=', ')))
```

If you look for their EnsemblID on Google they do come up, so I'm going to annotate them manually with their Ensembl ID and their biotype
```{r manually_annotate_missing_genes}
# Manually assign Ensembl IDs to the 15 genes
SFARI_genes = SFARI_genes %>% mutate(ID = case_when(`gene-symbol`=='MSNP1AS'        ~ 'ENSG00000251593',
                                                    `gene-symbol`=='RP11-1407O15.2' ~ 'ENSG00000174093',
                                                     TRUE ~ ID),
                                     gene_biotype = case_when(`gene-symbol`=='MSNP1AS'        ~ 'processed_pseudogene', 
                                                              `gene-symbol`=='RP11-1407O15.2' ~ 'protein_coding',
                                                               TRUE ~ gene_biotype))

# Check that all entries are complete
print(paste0('In the end, ', sum(is.na(SFARI_genes$gene_biotype)),' SFARI gene(s) don\'t have their biotype information',
      paste(SFARI_genes$`gene-symbol`[is.na(SFARI_genes$gene_biotype)], collapse = ', ')))
```

<br>

### Multiple Ensembl IDs for one gene

<br>

The relation between Ensembl IDs and genes should be 1:1, which is not the case here

*I think this isn't right, I think when the information of a gene gets updated it gets assigned a new Ensembl ID, so if you query different versions of BioMart, you will get many Ensembl IDs for a single gene
```{r}
print(paste0('There are ', length(unique(SFARI_genes$ID)), ' unique Ensembl IDs and ',
             length(unique(SFARI_genes$`gene-symbol`)),' unique gene symbols'))
```

<br>

All genes with more than one Ensembl ID assigned to them:

- For genes with two Ensembl IDs:

  - The smaller Ensembl ID (not starting with 2) is current and the larger one (starting with 2) is an older version for genes *ACTN4, ADORA3, APBA2, ARGHEF10, ARGGAP11B, ATXN7, CC2D1A, CLN8, CLSTN3, CNTNAP2, DEAF1, DNAH10, EPHB6, FAN1, HRAS*
  
  - In the previous version with the archive BioMart search, the larger Ensembl ID was current and the smaller one was an older ID for gene *EPPK1* (both IDs start with 2, so maybe it had something to do with that)
  
  - Both entries are current for genes *AFF2, AGTR, APH1A, ARHGEF9, ATP1A3, ATRX*
  
  - None of the entries are current for genes *CYFIB1, GGNBP2*

<br>

- For genes with more than two Ensembl IDs:

  - For *C4B*, all the Ensembl IDs are current, but all map to protein coding genes except for the 2nd one, which is a lncRNA. I think that they map to different "subtypes" of C4B, like C4B-202, C4B-217, C4B-201, and C4B-212
  
  - For HLA-A and HLA-B all the Ensembl IDs are current, I think they map to different subtypes of the gene, as above, but this time they all share the same gene_biotype
  
  - For *CNOT3* and *DLGAP2* only the first Ensembl ID was a current one, all the ones starting with 2 were old
  
<br>  
  
In conclusion, most of the times a duplicated Ensembl ID corresponds to an older Ensembl version, although not always, sometimes they map to different "versions" of the gene. I think it's not that bad because when crossing them with the Expression datasets, all the duplicates will disappear and this way we maximise the number of matches between datasets.

```{r genes_with_multiple_ensembl_ids}
duplicated_genes = SFARI_genes$`gene-symbol`[duplicated(SFARI_genes$`gene-symbol`)]

print(paste0('There are ', length(duplicated_genes), ' genes with multiple Ensembl IDs'))

SFARI_genes %>% dplyr::select(ID, `gene-symbol`, `gene-score`, syndromic, gene_biotype) %>% 
      filter(`gene-symbol` %in% duplicated_genes) %>% arrange(`gene-symbol`, ID) %>% kable(caption='Information of genes with more than one Ensembl ID')
```

<br>

### Gene Biotype

<br>

[Here](https://www.ensembl.org/info/genome/genebuild/biotypes.html) is the list of Ensembl Biotypes
```{r gene_biotypes}
table_info = SFARI_genes %>% apply_labels(gene_biotype = 'Gene Biotype')
cro(table_info$gene_biotype)
rm(table_info)
```

In the archive version of BioMart there's a biotype called processed transcript which in the newer version sometimes was labelled as protein_coding. The only two remaining occurrences of this have a new Ensembl ID with them, so we just have to be careful when labelling the transcriptomic data to update the gene_biotype to the newest version to avoid this.

```{r processed_transcript_biotype}
processed_transcripts = SFARI_genes$`gene-symbol`[SFARI_genes$gene_biotype=='processed_transcript']
SFARI_genes %>% filter(`gene-symbol` %in% processed_transcripts) %>% 
                dplyr::select(ID, `gene-symbol`, `gene-score`, syndromic, gene_biotype) %>% kable
```

<br>

### Write results to .csv file
```{r write_to_file}
write.csv(SFARI_genes, './../Data/SFARI_genes_08-29-2019_w_ensembl_IDs.csv', row.names=FALSE)
```

<br><br>

---

#### Session info
```{r print_session_info}
sessionInfo()
```
<br><br>
