---
title: 'Data Preprocessing'
output:
  html_document:
    code_folding: 'hide'
---

```{r load_packages, message=FALSE, warning=FALSE}

library(tidyverse) ; library(reshape2) ; library(glue) ; library(plotly) ; library(plotlyutils)
library(RColorBrewer) ; library(viridis) ; require(gridExtra) ; library(GGally) ; library(ggExtra) ; library(ggpubr)
library(biomaRt) ; library(DESeq2) ; library(sva) ; library(WGCNA) ; library(vsn)
library(dendextend) ; library(expss)
library(knitr) ; library(kableExtra)

```

---

## Raw data
<br><br>

Dataset downloaded from [Arkinglab website](http://www.arkinglab.org/resources/) in the *Transcriptome analysis reveals dysregulation of innate immune response genes and neuronal activity-dependent genes in autism* section.

### Load and annotate data

```{r load_and_annotate_data, message=FALSE, warning=FALSE}

# Load csvs
datExpr = read.delim('./../Data/datExpr.csv')
datMeta = read.delim('./../Data/datPheno.csv')

# Create dataset with gene information
datGenes = data.frame('ensembl_gene_id' = substr(datExpr$Gene, 1, 15), 
                      'hgnc_symbol' = substring(datExpr$Gene, 17))
rownames(datExpr) = datGenes$ensembl_gene_id
datExpr$Gene = NULL


### CLEAN METADATA DATA FRAME
datMeta = datMeta %>% dplyr::select('ID', 'case', 'sampleid', 'brainregion', 'positiononplate', 
                                    'Gender', 'Age', 'SiteHM', 'RIN', 'PMI', 'Dx') %>%
          mutate(brainregion = substr(ID,1,4)) %>%
          mutate(Brain_lobe = ifelse(brainregion=='ba19', 'Occipital', 'Frontal'),
                 Diagnosis = ifelse(Dx=='Autism', 'ASD', 'CTL')) %>%
          mutate(Diagnosis = factor(Diagnosis, levels = c('CTL','ASD'))) %>%
          dplyr::rename(Subject_ID = sampleid, Sex = Gender)



# GO Neuronal annotations: regex 'neuron' in GO functional annotations and label the genes that make a match as neuronal
GO_annotations = read.csv('./../Data/genes_GO_annotations.csv')
GO_neuronal = GO_annotations %>% filter(grepl('neuron', go_term)) %>% 
              mutate('ID'=as.character(ensembl_gene_id)) %>% 
              dplyr::select(-ensembl_gene_id) %>% distinct(ID) %>%
              mutate('Neuronal'=1)


# SFARI Genes
SFARI_genes = read_csv('./../../../SFARI/Data/SFARI_genes_01-03-2020_w_ensembl_IDs.csv')
SFARI_genes = SFARI_genes[!duplicated(SFARI_genes$ID) & !is.na(SFARI_genes$ID),]


# NCBI biotype annotation
NCBI_biotype = read.csv('./../../../NCBI/Data/gene_biotype_info.csv') %>% 
               dplyr::rename('ensembl_gene_id'=Ensembl_gene_identifier, 'gene_biotype'=type_of_gene, 
                             'hgnc_symbol'=Symbol) %>% 
               mutate(gene_biotype = ifelse(gene_biotype=='protein-coding','protein_coding', gene_biotype))


gg_colour_hue = function(n) {
  hues = seq(15, 375, length = n+1)
  pal = hcl(h = hues, l = 65, c = 100)[1:n]
}

rm(GO_annotations)
```

### Check sample composition
<br><br>

Data description taken from the paper [Transcriptome analysis reveals dysregulation of innate immune response genes and neuronal activity-dependent genes in autism](https://www.nature.com/articles/ncomms6748): 

Transcriptomes from 104 human brain cortical tissue samples were resolved using next-generation RNA sequencing technology at single-gene resolution and through co-expressing gene clusters or modules. Multiple cortical tissues corresponding to Brodmann Area 19 (BA19), Brodmann Area 10 (BA10) and Brodmann Area 44 (BA44) were sequenced in 62, 14 and 28 samples, respectively, resulting in a total of 57 (40 unique individuals) control and 47 (32 unique individuals) autism samples.

Brain tissue: Frozen brain samples were acquired through the Autism Tissue Program, with samples originating from two different sites: the Harvard Brain Tissue Resource Center and the NICHD Brain and Tissue Bank at the University of Maryland (Gandal's data were obtained also from the Autism Tissue Program, specifically from the Harvard Brain Bank).

Sequenced using Illumina’s HiSeq 2000 (Gandal used Illumina HiSeq 2500, they are compatible).

<br>
The dataset includes `r nrow(datExpr)` genes from `r  ncol(datExpr)` samples belonging to `r length(unique(datMeta$Subject_ID))` different subjects.
<br>

In the paper they talk about an original number of 110 samples and dropping 6 because of low gene coverage, resulting in 104 samples (which are the ones that are included in datMeta), but the expression dataset has 120 samples.

```{r}

no_metadata_samples = colnames(datExpr)[! colnames(datExpr) %in% datMeta$ID]
no_metadata_subjects = unique(substring(no_metadata_samples, 6))

```

Samples without Metadata information: `r paste(no_metadata_samples, collapse=', ')`
<br>

Samples without metadata but with a Subject ID that is included in the Metadata data frame `r paste(no_metadata_subjects[no_metadata_subjects %in% datMeta$Subject_ID], collapse=', ')`
<br>

Since we need the Phenotype information of the samples, I'm going to keep the samples that I can map to an existing Subject ID in the metadata and remove the others

```{r}
add_metadata_subjects = no_metadata_subjects[no_metadata_subjects %in% datMeta$Subject_ID]
add_metadata_samples = no_metadata_samples[grepl(paste(add_metadata_subjects, collapse='|'),
                                                 no_metadata_samples)]

for(sample in add_metadata_samples){
  new_row = datMeta %>% filter(Subject_ID == strsplit(sample,'\\.')[[1]][2]) %>% dplyr::slice(1) %>%
            mutate(ID = sample, brainregion = strsplit(sample,'\\.')[[1]][1],
                   Brain_lobe = ifelse(strsplit(sample,'\\.')[[1]][1]=='ba19','Occipital','Frontal'))
  datMeta = rbind(datMeta, new_row)
}

keep = substring(colnames(datExpr), 6) %in% datMeta$Subject_ID

datExpr = datExpr[,keep]

# Match order of datExpr columns and datMeta rows
datMeta = datMeta[match(colnames(datExpr), datMeta$ID),]

# Check they are in the same order
if(!all(colnames(datExpr) == datMeta$ID)){
  cat('\norder of samples don\'t match between datExpr and datMeta!\n')
}

rm(no_metadata_subjects, no_metadata_samples, add_metadata_subjects, add_metadata_samples, sample, new_row)
```

Removing `r sum(!keep)` samples (`r paste(colnames(datExpr)[!keep], collapse=', ')`) belonging to subjects with ID `r paste0(unique(substring(colnames(datExpr)[!keep],6)), collapse=', ')`

<br>
The dataset now has `r  ncol(datExpr)` samples belonging to `r length(unique(datMeta$Subject_ID))` different subjects.
<br>

```{r}
rm(keep)
```

**Counts distribution:** More than half of the counts are zero and most of the counts are relatively low, but there are some very high outliers
```{r, warning=FALSE, message=FALSE}

counts = datExpr %>% melt

count_distr = data.frame('Statistic' = c('Min', '1st Quartile', 'Median', 'Mean', '3rd Quartile', 'Max'),
                         'Values' = c(min(counts$value), quantile(counts$value, probs = c(.25, .5)) %>% unname,
                                      mean(counts$value), quantile(counts$value, probs = c(.75)) %>% unname,
                                      max(counts$value)))

count_distr %>% kable(digits = 2, format.args = list(scientific = FALSE)) %>% kable_styling(full_width = F)


rm(counts, count_distr)
```
<br>

**Diagnosis distribution by Sample:** There are a few more CTL samples than ASD
```{r}
table_info = datMeta %>% apply_labels(Diagnosis = 'Diagnosis', Brain_lobe = 'Brain Lobe', 
                                      SiteHM = 'Processing Group', Sex = 'Gender')
cro(table_info$Diagnosis)
```
<br>

**Diagnosis distribution by Subject:** There are more control subjects than ASD ones
```{r}
cro(table_info$Diagnosis[!duplicated(table_info$Subject_ID)])
```


**Brain region distribution:**  The Occipital lobe has more samples than the Frontal lobe, even though we are combining two brain regions in the Frontal Lobe
```{r}
cro(table_info$Brain_lobe)
```
<br>

Most of the Control samples are from the Occipital lobe, the Autism samples are balanced between both lobes. **This could cause problems because Control and Occipital are related**
```{r}
cro(table_info$Diagnosis, list(table_info$Brain_lobe,total()))
```
<br>

**Gender distribution:** There are thrice as many Male samples than Female ones
```{r}
cro(table_info$Sex)
```
<br>

There is a small imbalance between gender and diagnosis with more males in the control group than in the autism group
```{r}
cro(table_info$Diagnosis, list(table_info$Sex, total()))
```
<br>

**Age distribution:** Subjects between 2 and 82 years old with a mean of 18 years old
```{r, fig.width=10}
summary(datMeta$Age)

datMeta_by_subject = datMeta %>% filter(!duplicated(Subject_ID))
datMeta_by_subject %>% ggplot(aes(Age)) +
                       geom_density(alpha=0.2, aes(group = Diagnosis, fill = Diagnosis, color = Diagnosis)) +
                       geom_density(alpha=0.3, fill='gray', color='gray') +
                       theme_minimal()


rm(datMeta_by_subject)
```
<br>

<br>

**Processing Group and Diagnosis:** There is a relation between processing group and Diagnosis! **This can be a problem because processing group is an important Batch Correction covariate, but if it's related to Diagnosis, we risk inadvertedly losing information related to ASD**
```{r}
cro(table_info$Diagnosis, list(table_info$SiteHM,total()))

rm(table_info)
```

<br><br>

### Annotate genes with BioMart information
<br><br>

I was originally running this with the feb2014 version of BioMart because that's the one that Gandal used (and it finds all of the Ensembl IDs, which other versions don't), but it has some outdated biotype annotations, to fix this I'll obtain all the information except the biotype label from BioMart in the same way as it had been done before, and then I'll add the most current biotype label using information from NCBI's website and then from BioMart in the following way:

1. Use BioMart to run a query with the original feb2014 version using the Ensembl IDs as keys to obtain all the information except the biotype labels (we are not going to retrieve the gene name from biomart because we already extracted it from datExpr)

2. Annotate genes with Biotype labels:

    2.1 Use the NCBI annotations downloaded from [NCBI's website](https://www.ncbi.nlm.nih.gov/home/download/) and processed in NCBI/RMarkdowns/clean_data.html (there is information for only 26K genes, so some genes will remain unlabelled)
  
    2.2 Use the current version (jan2020) to obtain the biotype annotations using the Ensembl ID as keys (some genes don't return a match)
  
    2.3 For the genes that didn't return a match, use the current version (jan2020) to obtain the biotype annotations using the gene name as keys
  
    2.4 For the genes that returned multiple labels, use the feb2014 version with the Ensembl IDs as keys

<br>

**Note:** A small proportion of genes don't make a match in any of these queries, so they will be lost when we start filtering out genes
<br>
```{r annotate_genes, warning=FALSE, message=FALSE, fig.height=2, fig.align='center'}

labels_source = data.frame('source' = c('NCBI', 'BioMart2020_byID', 'BioMart2020_byGene', 'BioMart2014'),
                                      'n_matches' = rep(0,4))

########################################################################################
# 1. Query archive version

getinfo = c('ensembl_gene_id','chromosome_name','start_position','end_position','strand')
mart = useMart(biomart = 'ENSEMBL_MART_ENSEMBL', dataset = 'hsapiens_gene_ensembl', 
               host = 'feb2014.archive.ensembl.org')
datGenes_BM = getBM(attributes = getinfo, filters = c('ensembl_gene_id'), values = rownames(datExpr), 
                    mart = mart)

datGenes = datGenes %>% left_join(datGenes_BM, by = 'ensembl_gene_id')

datGenes$length = datGenes$end_position - datGenes$start_position

cat(paste0('1. ', sum(is.na(datGenes$start_position)), '/', nrow(datGenes),
             ' Ensembl IDs weren\'t found in the feb2014 version of BioMart'))


########################################################################################
########################################################################################
# 2. Get Biotype Labels

cat('2. Add biotype information')

########################################################################################
# 2.1 Add NCBI annotations
datGenes = datGenes %>% left_join(NCBI_biotype, by=c('ensembl_gene_id','hgnc_symbol'))

cat(paste0('2.1 ' , sum(is.na(datGenes$gene_biotype)), '/', nrow(datGenes),
             ' Ensembl IDs weren\'t found in the NCBI database'))

labels_source$n_matches[1] = sum(!is.na(datGenes$gene_biotype))

########################################################################################
# 2.2 Query current BioMart version for gene_biotype using Ensembl ID as key

getinfo = c('ensembl_gene_id','gene_biotype')
mart = useMart(biomart = 'ENSEMBL_MART_ENSEMBL', dataset = 'hsapiens_gene_ensembl',
               host = 'jan2020.archive.ensembl.org')
datGenes_biotype = getBM(attributes = getinfo, filters = c('ensembl_gene_id'), mart=mart, 
                         values = datGenes$ensembl_gene_id[is.na(datGenes$gene_biotype)])

cat(paste0('2.2 ' , sum(is.na(datGenes$gene_biotype))-nrow(datGenes_biotype), '/', 
           sum(is.na(datGenes$gene_biotype)),
           ' Ensembl IDs weren\'t found in the jan2020 version of BioMart when querying by Ensembl ID'))

# Add new gene_biotype info to datGenes
datGenes = datGenes %>% left_join(datGenes_biotype, by='ensembl_gene_id') %>%
           mutate(gene_biotype = coalesce(as.character(gene_biotype.x), gene_biotype.y)) %>%
           dplyr::select(-gene_biotype.x, -gene_biotype.y)

labels_source$n_matches[2] = sum(!is.na(datGenes$gene_biotype)) - labels_source$n_matches[1]

########################################################################################
# 3. Query current BioMart version for gene_biotype using gene symbol as key

missing_genes = unique(datGenes$hgnc_symbol[is.na(datGenes$gene_biotype)])
getinfo = c('hgnc_symbol','gene_biotype')
datGenes_biotype_by_gene = getBM(attributes = getinfo, filters = c('hgnc_symbol'), mart = mart,
                                 values = missing_genes)

cat(paste0('2.3 ', length(missing_genes)-length(unique(datGenes_biotype_by_gene$hgnc_symbol)),'/',
           length(missing_genes),
           ' genes weren\'t found in the current BioMart version when querying by gene name'))

dups = unique(datGenes_biotype_by_gene$hgnc_symbol[duplicated(datGenes_biotype_by_gene$hgnc_symbol)])
cat(paste0('    ', length(dups), ' genes returned multiple labels (these won\'t be added)'))

# Update information
datGenes_biotype_by_gene = datGenes_biotype_by_gene %>% filter(!hgnc_symbol %in% dups)
datGenes = datGenes %>% left_join(datGenes_biotype_by_gene, by='hgnc_symbol') %>% 
           mutate(gene_biotype = coalesce(gene_biotype.x, gene_biotype.y)) %>%
           dplyr::select(-gene_biotype.x, -gene_biotype.y)

labels_source$n_matches[3] = sum(!is.na(datGenes$gene_biotype)) - sum(labels_source$n_matches)

########################################################################################
# 4. Query feb2014 BioMart version for the missing biotypes

missing_ensembl_ids = unique(datGenes$ensembl_gene_id[is.na(datGenes$gene_biotype)])

getinfo = c('ensembl_gene_id','gene_biotype')
mart = useMart(biomart = 'ENSEMBL_MART_ENSEMBL', dataset = 'hsapiens_gene_ensembl', 
               host = 'feb2014.archive.ensembl.org')
datGenes_biotype_archive = getBM(attributes = getinfo, filters=c('ensembl_gene_id'), 
                                 values = missing_ensembl_ids, mart=mart)

cat(paste0('2.4 ', length(missing_ensembl_ids)-nrow(datGenes_biotype_archive),'/',length(missing_ensembl_ids),
             ' genes weren\'t found in the feb2014 BioMart version when querying by Ensembl ID'))

# Update information
datGenes = datGenes %>% left_join(datGenes_biotype_archive, by='ensembl_gene_id') %>% 
            mutate(gene_biotype = coalesce(gene_biotype.x, gene_biotype.y)) %>%
            dplyr::select(-gene_biotype.x, -gene_biotype.y)

labels_source$n_matches[4] = sum(!is.na(datGenes$gene_biotype)) - sum(labels_source$n_matches)

########################################################################################
# Plot results

labels_source = labels_source %>% add_row(source = 'missing', 
                                          n_matches = nrow(datGenes) - sum(labels_source$n_matches)) %>% 
                mutate(x = 1, percentage = round(100*n_matches/sum(n_matches),2),
                       source = factor(source, levels=c('BioMart2014','BioMart2020_byGene','BioMart2020_byID',
                                                        'NCBI','missing')))
                

p = labels_source %>% ggplot(aes(x, percentage, fill=source)) + geom_bar(position='stack', stat='identity') +
    theme_minimal() + coord_flip() + theme(legend.position='bottom', axis.title.y=element_blank(),
    axis.text.y=element_blank(), axis.ticks.y=element_blank()) + ylab('Percentage of genes') +
    scale_fill_manual(values = c(gg_colour_hue(nrow(labels_source)-1),'gray'))

ggplotly(p + theme(legend.position='none'))
as_ggplot(get_legend(p))

########################################################################################
# Reorder rows to match datExpr
datGenes = datGenes[match(rownames(datExpr), datGenes$ensembl_gene_id),]


rm(getinfo, mart, datGenes_BM, datGenes_biotype, datGenes_biotype_by_gene, datGenes_biotype_archive,
   dups, missing_ensembl_ids, missing_genes, labels_source, p)
```

---

## Filtering
<br><br>

Checking how many SFARI genes are in the dataset

```{r}
df = SFARI_genes %>% dplyr::select(-gene_biotype) %>% inner_join(datGenes, by=c('ID'='ensembl_gene_id'))
n_SFARI = df[['gene-symbol']] %>% unique %>% length
```

Considering all genes, this dataset contains `r df[['gene-symbol']] %>% unique %>% length` of the `r SFARI_genes[['gene-symbol']] %>% unique %>% length` SFARI genes
<br><br>

**1.-** Filter entries for which we didn't manage to obtain its genotype information
<br>

1.1 Missing Biotype
```{r filter_genes_wo_length}

to_keep = !is.na(datGenes$gene_biotype)

datGenes = datGenes[to_keep,]
datExpr = datExpr[to_keep,]
rownames(datGenes) = datGenes$ensembl_gene_id

```

Removed `r sum(!to_keep)` 'genes', `r sum(to_keep)` remaining
<br>

Filtering genes without biotype information, we are left with `r df[['gene-symbol']][!is.na(df$gene_biotype)] %>% unique %>% length` SFARI Genes (we lose `r n_SFARI - df[['gene-symbol']][!is.na(df$gene_biotype)] %>% unique %>% length` genes)

<br><br><br>

1.2 Missing Length of the sequence
```{r}

to_keep = !is.na(datGenes$length)

datExpr = datExpr[to_keep,]
datGenes = datGenes[to_keep,]

```

Removed `r sum(!to_keep)` 'genes', `r sum(to_keep)` remaining
<br>

Filtering genes without sequence length information, we are left with `r df[['gene-symbol']][!is.na(df$gene_biotype)] %>% unique %>% length` SFARI Genes (we lose `r n_SFARI - df[['gene-symbol']][!is.na(df$gene_biotype)] %>% unique %>% length` genes)

<br><br><br>

**2.-** Filter genes that do not encode any protein

<br>

`r round(100*mean(datGenes$gene_biotype == 'protein_coding'))`% of the genes are protein coding genes

```{r gene_biotype_table}
datGenes$gene_biotype %>% table %>% sort(decreasing=TRUE) %>% kable(caption='Biotypes of genes in dataset') %>%
                          kable_styling(full_width = F)
```

Most of the non-protein coding genes have very low levels of expression
```{r explore_non_protein_coding_genes, fig.width=10}
plot_data = data.frame('ID' = rownames(datExpr), 'MeanExpr' = apply(datExpr, 1, mean),
                       'ProteinCoding' = datGenes$gene_biotype=='protein_coding')

ggplotly(plot_data %>% ggplot(aes(log2(MeanExpr+1), fill=ProteinCoding, color=ProteinCoding)) + 
         geom_density(alpha=0.5) + theme_minimal())


rm(plot_data)
```


```{r protein_coding_genes_and_SFARI_score}

df = SFARI_genes %>% dplyr::select(-gene_biotype) %>% inner_join(datGenes, by=c('ID'='ensembl_gene_id'))

```

Filtering protein coding genes, we are left with `r df[['gene-symbol']][df$gene_biotype=='protein_coding'] %>% unique %>% length` SFARI Genes (we lose `r n_SFARI - df[['gene-symbol']][df$gene_biotype=='protein_coding'] %>% unique %>% length` genes)
<br>

**Note:** The gene name for Ensembl ID ENSG00000187951 is wrong, it should be AC091057.1 instead of ARHGAP11B, but the biotype is right, so it would still be filtered out

```{r}
n_SFARI = df[['gene-symbol']][df$gene_biotype=='protein_coding'] %>% unique %>% length

df %>% filter(!`gene-symbol` %in% df$`gene-symbol`[df$gene_biotype=='protein_coding']) %>% 
       dplyr::select(ID, `gene-symbol`, `gene-score`, gene_biotype, syndromic, `number-of-reports`) %>% 
       kable(caption='Lost Genes')  %>% kable_styling(full_width = F)

rm(df)
```

```{r filter_non_protein_coding_genes}

if(!all(rownames(datExpr)==rownames(datGenes))) cat('!!! gene rownames do not match!!!')

to_keep = datGenes$gene_biotype == 'protein_coding'
datExpr = datExpr %>% filter(to_keep)
datGenes = datGenes %>% filter(to_keep)
rownames(datExpr) = datGenes$ensembl_gene_id
rownames(datGenes) = datGenes$ensembl_gene_id

```
<br>

Removed `r sum(!to_keep)` genes. `r sum(to_keep)` remaining


<br>
**3.-** Filter genes with low expression levels

$\qquad$ 3.1 Remove genes with zero expression in all of the samples
```{r, warning=FALSE}

to_keep = rowSums(datExpr) > 0

df = data.frame('rowSums' = rowSums(datExpr), 'ensembl_gene_id' = rownames(datExpr)) %>%
     right_join(SFARI_genes, by='ensembl_gene_id') %>% filter(rowSums == 0 & !is.na(`gene-score`)) %>%
     arrange(`gene-score`) %>% dplyr::select(-ensembl_gene_id) %>%
     filter(!duplicated(`gene-symbol`), !`gene-symbol` %in% datGenes$hgnc_symbol[to_keep])

datExpr = datExpr[to_keep,]
datGenes = datGenes[to_keep,]

```
<br>

Removed `r sum(!to_keep)` genes. `r sum(to_keep)` remaining
<br>

`r SFARI_genes[['gene-symbol']][SFARI_genes$ID %in% rownames(datExpr)] %>% unique %>% length` SFARI genes remaining (we lost `r n_SFARI - SFARI_genes[['gene-symbol']][SFARI_genes$ID %in% rownames(datExpr)] %>% unique %>% length` genes)

```{r}

df %>% dplyr::select(ID, `gene-symbol`, `gene-score`, gene_biotype, syndromic, `number-of-reports`) %>% 
       kable(caption='Lost SFARI Genes') %>% kable_styling(full_width = F)

n_SFARI = SFARI_genes[['gene-symbol']][SFARI_genes$ID %in% rownames(datExpr)] %>% unique %>% length


rm(df)
```

<br>

$\qquad$ 3.2 Removing genes with a high percentage of zeros

<br>

Choosing the threshold:

**Criteria for selecting the percentage of zeros threshold:** The minimum value in which the preprocessed data is relatively homoscedastic (we're trying to get rid of the group of genes with very low mean and SD that make the cloud of points look like a comic book speech bubble)

- On the plot I'm using the "dual" of the maximum percentage of zeros, the minimum percentage of non zeros so the visualisation is more intuitive

- 75% seems to be a good threshold for the minimum percentage of non zeros, so 25% will be the maximum percentage of zeros allowed in a row

- The Mean vs SD plot doesn't show all of the genes, a random sample was selected for the genes with higher level of expression so the visualisation wouldn't be as heavy (and since we care about the genes with the lowest levels of expression, we aren't losing important information)

```{r}
datMeta_original = datMeta
datExpr_original = datExpr
datGenes_original = datGenes
```

```{r warning=FALSE, message=FALSE, fig.width=10}
thresholds = round(100*(ncol(datExpr)-c(seq(100,7,-10),5,3,2,0))/ncol(datExpr))

for(threshold in thresholds){
  
  datMeta = datMeta_original
  datExpr = datExpr_original
  datGenes = datGenes_original
  
  to_keep = apply(datExpr, 1, function(x) 100*mean(x>0)) >= threshold
  datGenes = datGenes[to_keep,]
  datExpr = datExpr[to_keep,]
  
  # Filter outlier samples
  absadj = datExpr %>% bicor %>% abs
  netsummary = fundamentalNetworkConcepts(absadj)
  ku = netsummary$Connectivity
  z.ku = (ku-mean(ku))/sqrt(var(ku))
  
  to_keep = z.ku > -2
  datMeta = datMeta[to_keep,]
  datExpr = datExpr[,to_keep]
  
  rm(absadj, netsummary, ku, z.ku, to_keep)
  
  
  # Create a DeseqDataSet object, estimate the library size correction and save the normalized counts matrix
  counts = datExpr %>% as.matrix
  rowRanges = GRanges(datGenes$chromosome_name,
                      IRanges(start=datGenes$start_position, width=datGenes$length),
                      strand=datGenes$strand,
                      feature_id=datGenes$ensembl_gene_id)
  se = SummarizedExperiment(assays=SimpleList(counts=counts), rowRanges=rowRanges, colData=datMeta)
  dds = DESeqDataSet(se, design =~Diagnosis)
  
  # Perform vst
  vsd = vst(dds)
  
  datExpr_vst = assay(vsd)
  datMeta_vst = colData(vsd)
  datGenes_vst = rowRanges(vsd)
  
  rm(counts, rowRanges, se, vsd)
  
  # Save summary results in dataframe
  if(threshold == thresholds[1]){
    mean_vs_sd_data = data.frame('threshold'=threshold, 'ID'=rownames(datExpr_vst),
                                 'Mean'=rowMeans(datExpr_vst), 'SD'=apply(datExpr_vst,1,sd))
  } else {
    new_entries = data.frame('threshold'=threshold, 'ID'=rownames(datExpr_vst),
                                 'Mean'=rowMeans(datExpr_vst), 'SD'=apply(datExpr_vst,1,sd))
    mean_vs_sd_data = rbind(mean_vs_sd_data, new_entries)
  }
}

# Visualise the effects of different thresholds
to_keep_1 = mean_vs_sd_data$ID[mean_vs_sd_data$threshold==thresholds[1] & mean_vs_sd_data$Mean<7] %>%
            as.character
to_keep_2 = mean_vs_sd_data$ID[mean_vs_sd_data$threshold==thresholds[1] & mean_vs_sd_data$Mean>=7]
to_keep_2 = to_keep_2 %>% sample(round(length(to_keep_2)/10)) %>% as.character

plot_data = mean_vs_sd_data[mean_vs_sd_data$ID %in% c(to_keep_1, to_keep_2),]

ggplotly(plot_data %>% ggplot(aes(Mean, SD)) + 
         geom_point(color='#0099cc', alpha=0.2, aes(id=ID, frame=threshold)) + 
         scale_x_log10() + scale_y_log10() + theme_minimal())

# Plot remaining genes
plot_data = mean_vs_sd_data %>% group_by(threshold) %>% tally

ggplotly(plot_data %>% ggplot(aes(threshold, n)) + geom_point() + geom_line() +
         theme_minimal() + ggtitle('Remaining genes for each filtering threshold'))


rm(to_keep_1, to_keep_2, plot_data, dds, thresholds)
```

```{r}
# Return to original variables
datExpr = datExpr_original
datGenes = datGenes_original
datMeta = datMeta_original

rm(datExpr_original, datGenes_original, datMeta_original, datExpr_vst, datGenes_vst, datMeta_vst)
```

<br>

Selecting a threshold of 85
```{r, warning=FALSE, fig.width=10}

# Minimum percentage of non-zero entries allowed per gene
threshold = 85

plot_data = data.frame('id'=rownames(datExpr),
                       'non_zero_percentage' = apply(datExpr, 1, function(x) 100*mean(x>0)))

ggplotly(plot_data %>% ggplot(aes(x=non_zero_percentage)) + 
         geom_density(color='#0099cc', fill='#0099cc', alpha=0.3) + 
         geom_vline(xintercept=threshold, color='gray') +
         xlab('% of non-zero entries') + ylab('Density') +
         ggtitle('Percentage of non-zero entries distribution') + theme_minimal())

to_keep = apply(datExpr, 1, function(x) 100*mean(x>0)) >= threshold
datGenes = datGenes[to_keep,]
datExpr = datExpr[to_keep,]

```

Removed `r sum(!to_keep)` genes. `r sum(to_keep)` remaining
<br>

`r SFARI_genes[['gene-symbol']][SFARI_genes$ID %in% rownames(datExpr)] %>% unique %>% length` SFARI genes remaining (we lost `r n_SFARI - SFARI_genes[['gene-symbol']][SFARI_genes$ID %in% rownames(datExpr)] %>% unique %>% length` genes)

```{r}
n_SFARI = SFARI_genes[['gene-symbol']][SFARI_genes$ID %in% rownames(datExpr)] %>% unique %>% length

rm(threshold, plot_data, to_keep)
```

<br>
**4.-** Filter outlier samples

Using node connectivity as a distance measure, normalising it and filtering out genes farther away than 2 standard deviations from the left (lower connectivity than average, not higher)

- Gandal uses the formula $s_{ij}=\frac{1+bw(i,j)}{2}$ to convert all the weights to positive values, but I used $s_{ij}=|bw(i,j)|$ instead because I think it makes more sense. In the end it doesn't matter because they select as outliers the same six samples

- Only 2 outliers, both with ASD Diagnosis

```{r, warning=FALSE}
absadj = datExpr %>% bicor %>% abs
netsummary = fundamentalNetworkConcepts(absadj)
ku = netsummary$Connectivity
z.ku = (ku-mean(ku))/sqrt(var(ku))

plot_data = data.frame('sample'=1:length(z.ku), 'distance'=z.ku, 'Sample_ID'=datMeta$ID, 
                       'Subject_ID'=datMeta$Subject_ID, 'Extraction_Batch'=datMeta$SiteHM,
                       'Brain_Lobe'=datMeta$Brain_lobe, 'Sex'=datMeta$Sex, 'Age'=datMeta$Age,
                       'Diagnosis'=datMeta$Diagnosis, 'PMI'=datMeta$PMI)

selectable_scatter_plot(plot_data, plot_data[,-c(1:3)])
```

Outlier samples: `r paste(as.character(plot_data$Sample_ID[plot_data$distance< -2]), collapse=', ')`

```{r}
to_keep = z.ku > -2
datMeta = datMeta[to_keep,]
datExpr = datExpr[,to_keep]

rm(absadj, netsummary, ku, z.ku, plot_data)
```

Removed `r sum(!to_keep)` samples, `r sum(to_keep)` remaining

```{r}
rm(to_keep)
```
<br><br><br>

<br>
**5.-** Filter repeated genes

There are `r nrow(datGenes) - datGenes$hgnc_symbol %>% unique %>% length` genes with more than one ensembl ID in the dataset. To accurately refer to the rows of my data as 'genes', I'm going to remove the repeated ones.

```{r}
dup_genes = datGenes$hgnc_symbol %>% duplicated

datGenes = datGenes[!dup_genes,]
datExpr = datExpr[!dup_genes,]
```

Removed `r sum(dup_genes)` genes. `r sum(!dup_genes)` remaining
<br>

`r SFARI_genes[['gene-symbol']][SFARI_genes$ID %in% rownames(datExpr)] %>% unique %>% length` SFARI genes remaining (we lost `r n_SFARI - SFARI_genes[['gene-symbol']][SFARI_genes$ID %in% rownames(datExpr)] %>% unique %>% length` genes)

```{r}
rm(dup_genes, n_SFARI)
```
<br>

**6.-** Filter Samples to correct Sample's Metadata imbalances related to Diagnosis
<br><br>

We have two varibles with imbalances related to Diagnosis: Brain Lobe and Processing Group, with the Occipital lobe and the Maryland processing group being biased towards the Control group. 

- Balancing the samples by Brain Lobe can be done relatively easy, by removing 9 Control samples from the Occipital lobe and 8 Autism samples from the Frontal lobe (this way we'll obtain a 2:3 ration between Frontal and Occipital lobe for both Diagnosis groups)

- Balancing the samples by Processing Group is a lot more complicated since would require to remove more than 50 samples (almost half of the samples), since Harvard has a ratio of 3:1 between ASD and Control samples and Maryland of 1:4, (in the opposite direction)

Because the cost of balancing the samples by Processing Group is too high, I'm not going to try to do it directly, but I'm going to chose the samples I'm going to remove to balance the samples by Brain Lobe considering the Processing site to at least reduce this imbalance a little

```{r}

set.seed(123)

ctl_samples = datMeta %>% filter(Diagnosis == 'CTL'& Brain_lobe == 'Occipital' & SiteHM == 'M') %>%
              sample_n(9) %>% pull(ID) %>% as.character

asd_samples = datMeta %>% filter(Diagnosis == 'ASD'& Brain_lobe == 'Frontal' & SiteHM == 'H') %>%
              sample_n(8) %>% pull(ID) %>% as.character

to_keep = !datMeta$ID %in% c(ctl_samples, asd_samples)

datMeta = datMeta[to_keep,]
datExpr = datExpr[,to_keep]

```

Removing the ASD samples with ID: `r paste(asd_samples, sep=', ')`
<br>

Removing the Control samples with ID: `r paste(ctl_samples, sep=', ')`
<br><br>

The samples are now balanced by Brain Lobe and they are still unbalanced by Processing Group, but at least not as much as before (Harvard 2:1 and Maryland 1:3)
```{r}
rm(ctl_samples, asd_samples, to_keep)

table_info = datMeta %>% apply_labels(Diagnosis = 'Diagnosis', Brain_lobe = 'Brain Lobe', 
                                      SiteHM = 'Processing Group')

cro(table_info$Diagnosis, list(table_info$Brain_lobe,total()))

cro(table_info$Diagnosis, list(table_info$SiteHM,total()))
```
<br>

After filtering, the dataset consists of `r nrow(datExpr)` genes and `r ncol(datExpr)` samples
<br><br>

### Save filtered and annotated dataset

```{r}
save(datExpr, datMeta, datGenes, file='./../Data/filtered_raw_data.RData')
#load('./../Data/filtered_raw_data.RData')
```

---

## Batch Effects Exploratory Analysis
<br><br>

**Note:** No batch correction is performed in this section, this is done after the normalisation step

According to [Tackling the widespread and critical impact of batch effects in high-throughput data](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC3880143/), technical artifacts can be an important source of variability in the data, so batch correction should be part of the standard preprocessing pipeline of gene expression data.

They say Processing group and Date of the experiment are good batch surrogates, I only have processing group, so I'm going to see if this affects the data in any clear way to use it as a surrogate.
<br>

### Processing group
<br>

Data was processed by two different groups: the Harvard Brain Tissue Resource Center (H) and the NICHD Brain and Tissue Bank at the University of Maryland (M)

```{r}
table_info = datMeta %>% apply_labels(SiteHM = 'Processing Group', Diagnosis = 'Diagnosis')

table_info$SiteHM %>% cro
```
<br>

Samples don’t seem to cluster together that strongly by batch, and in the cases that they do align, since we know there is a relation between Processing Group and Diagnosis, we don't which of these variables is responsible for it

```{r samples_histogram, fig.width=10}
h_clusts = datExpr %>% t %>% dist %>% hclust %>% as.dendrogram

create_viridis_dict = function(){
  min_age = datMeta$Age %>% min
  max_age = datMeta$Age %>% max
  viridis_age_cols = viridis(max_age - min_age + 1)
  names(viridis_age_cols) = seq(min_age, max_age)
  
  return(viridis_age_cols)
}
viridis_age_cols = create_viridis_dict()

dend_meta = datMeta[match(labels(h_clusts), datMeta$ID),] %>% 
            mutate('Site' = ifelse(SiteHM=='H', '#F8766D', '#00BFC4'),
                   'Diagnosis' = ifelse(Diagnosis=='CTL','#008080','#86b300'), # Blue control, Green ASD
                   'Sex' = ifelse(Sex=='F','#ff6666','#008ae6'),            # Pink Female, Blue Male
                   'Region' = case_when(Brain_lobe=='Frontal'~'#F8766D',        # ggplot defaults for 2 colours
                                        Brain_lobe=='Occipital'~'#00BFC4'),
                   'Age' = viridis_age_cols[as.character(Age)]) %>%             # Purple: young, Yellow: old
            dplyr::select(Age, Region, Sex, Diagnosis, Site)
h_clusts %>% dendextend::set('labels', rep('', nrow(datMeta))) %>% 
             dendextend::set('branches_k_color', k=9) %>% plot
colored_bars(colors=dend_meta)

rm(h_clusts, dend_meta, create_viridis_dict, viridis_age_cols)
```

- Comparing the mean expression of each sample by batch we can see there is some batch effect differentiating them

- We were expecting some kind of imbalance here since we know that ASD samples have a higher general level of expression when compared to the Control group

- What we weren't expecting to find is the Maryland group to have a higher level of expression than the Harvard group (since this group has more Control samples than ASD and the Harvard group has the opposite bias)

```{r, warning=FALSE, message=FALSE}
plot_data_b1 = data.frame('Mean'=colMeans(datExpr[,datMeta$SiteHM=='H']), 'Batch'='H')
plot_data_b2 = data.frame('Mean'=colMeans(datExpr[,datMeta$SiteHM=='M']), 'Batch'='M')

plot_data = rbind(plot_data_b1, plot_data_b2)
mu = plot_data %>% group_by(Batch) %>% dplyr::summarise(BatchMean=mean(Mean))

ggplotly(plot_data %>% ggplot(aes(x=Mean, color=Batch, fill=Batch)) + geom_density(alpha=0.3) + 
         geom_vline(data=mu, aes(xintercept=BatchMean, color=Batch), linetype='dashed') +
         ggtitle('Mean expression by sample grouped by Batch') + scale_x_log10() + theme_minimal())

rm(plot_data_b1, plot_data_b2, plot_data, mu)
```
<br><br>

### Looking for unknown sources of batch effects
<br>

Following the pipeline from [Surrogate variable analysis: hidden batch effects](https://biodatascience.github.io/compbio/dist/sva.html) where sva is used with DESeq2:
<br>

Create a DeseqDataSet object, estimate the library size correction and save the normalized counts matrix
```{r}
counts = datExpr %>% as.matrix
rowRanges = GRanges(datGenes$chromosome_name,
                  IRanges(datGenes$start_position, width=datGenes$length),
                  strand=datGenes$strand,
                  feature_id=datGenes$ensembl_gene_id)
se = SummarizedExperiment(assays=SimpleList(counts=counts), rowRanges=rowRanges, colData=datMeta)
dds = DESeqDataSet(se, design = ~ Diagnosis)

dds = estimateSizeFactors(dds)
norm.cts = counts(dds, normalized = TRUE)
```

Provide the normalized counts and two model matrices to SVA. The first matrix uses the biological condition, and the second model matrix is the null model.
```{r}
mod = model.matrix(~ Diagnosis, colData(dds))
mod0 = model.matrix(~ 1, colData(dds))
sva_fit = svaseq(norm.cts, mod=mod, mod0=mod0)

rm(mod, mod0, norm.cts)
```

Found `r ncol(sva_fit$sv)` surrogate variables, since there is no direct way to select which ones to pick [Bioconductor answer](https://support.bioconductor.org/p/71447/), kept all of them.

Include SV estimations to datMeta information
```{r}
sv_data = sva_fit$sv %>% data.frame
colnames(sv_data) = paste0('SV', 1:ncol(sv_data))

datMeta_sva = cbind(datMeta, sv_data)

rm(sv_data, sva_fit)
```

**In conclusion:** Site could work as a surrogate for batch effects, but has the huge downside that is correlated to Diagnosis! The sva package found other 23 variables that could work as surrogates which are now included in datMeta and should be included in the DEA.
<br><br>

<br><br>

---

## Normalisation and Differential Expression Analysis
<br><br>

Using DESeq2 package to perform normalisation. I chose this package over limma because limma uses the log transformed data as input instead of the raw counts and I have discovered that in this dataset, this transformation affects genes differently depending on their mean expression level, and genes with a high SFARI score are specially affected by this.

```{r}
plot_data = data.frame('ID'=rownames(datExpr), 'Mean'=rowMeans(datExpr), 'SD'=apply(datExpr,1,sd))

plot_data %>% ggplot(aes(Mean, SD)) + geom_point(color='#0099cc', alpha=0.1) + geom_abline(color='gray') +
              scale_x_log10() + scale_y_log10() + theme_minimal()

rm(plot_data)
```

- **Using vst** instead of rlog to perform normalisation. [Bioconductor question](https://support.bioconductor.org/p/104615/) explaining differences between methods. Chose vst because **a)** it is much faster than rlog (it is recommended to use vst for samples larger than 50), and **b)** Michael Love (author of DESEq2) recommends using it over rlog

- **Including a log fold change threshold of 0 in the results formula** $H_0:lfc=0$ because setting any other log fold change seems arbitrary and we risk losing genes with a significant differential expression for genes with a higher fold change, but not necessarily as significant.

```{r normalisation}
counts = datExpr %>% as.matrix
rowRanges = GRanges(datGenes$chromosome_name,
                  IRanges(datGenes$start_position, width=datGenes$length),
                  strand=datGenes$strand,
                  feature_id=datGenes$ensembl_gene_id)
se = SummarizedExperiment(assays=SimpleList(counts=counts), rowRanges=rowRanges, colData=datMeta_sva)
dds = DESeqDataSet(se, design = ~ SiteHM + SV1 + SV2 + SV3 + SV4 + SV5 + SV6 + SV7 + SV8 + SV9 + 
                                  SV10 + SV11 + SV12 + SV13 + SV14 + SV15 + SV16 + SV17 + SV18 +
                                  SV19 + SV20 + Diagnosis)

# Perform DEA
#dds = DESeq(dds) # Changed this for the three lines below because some rows don't converge
dds = estimateSizeFactors(dds)
dds = estimateDispersions(dds)
dds = nbinomWaldTest(dds, maxit=10000)

DE_info = results(dds, lfcThreshold=0, altHypothesis='greaterAbs')

# Perform vst
vsd = vst(dds)

datExpr_vst = assay(vsd)
datMeta_vst = colData(vsd)
datGenes_vst = rowRanges(vsd)

rm(counts, rowRanges, se, vsd)
```

Using the plotting function [DESEq2's manual](http://bioconductor.org/packages/devel/bioc/vignettes/DESeq2/inst/doc/DESeq2.html) proposes to study vst's output it looks like the data could be homoscedastic
```{r}
meanSdPlot(datExpr_vst, plot=FALSE)$gg + theme_minimal()
```

Plotting points individually we can notice some heteroscedasticity in the data

```{r warning=FALSE, message=FALSE}
plot_data = data.frame('ID'=rownames(datExpr_vst), 'Mean'=rowMeans(datExpr_vst), 'SD'=apply(datExpr_vst,1,sd))

plot_data %>% ggplot(aes(Mean, SD)) + geom_point(color='#0099cc', alpha=0.2) + geom_smooth(color = 'gray') +
              scale_x_log10() + scale_y_log10() + theme_minimal()

rm(plot_data)
```
<br><br>

Rename normalised datasets to continue working with these
```{r}
datExpr = datExpr_vst
datMeta = datMeta_vst %>% data.frame
datGenes = datGenes_vst

rm(datExpr_vst, datMeta_vst, datGenes_vst, datMeta_sva)
```

<br><br>

---

## Batch Effect Correction
<br><br>

By including the surrogate variables in the DESeq formula we only modelled the batch effects into the DEA, but we didn't actually correct them from the data, for that we need to use ComBat (or other equivalent package) in the already normalised data
<br>

### SVA surrogate variables
<br>

In some places they say you shouldn't correct these effects on the data because you risk losing biological variation, in others they say you should because they introduce noise to the data. The only thing everyone agrees on is that you shouldn't remove them before performing DEA but instead include them in the model.

Based on the conclusions from [Practical impacts of genomic data “cleaning” on biological discovery using surrogate variable analysis](https://bmcbioinformatics.biomedcentral.com/track/pdf/10.1186/s12859-015-0808-5) it seems like it may be a good idea to remove the batch effects from the data and not only from the DE analysis:

- Using SVA, ComBat or related tools can increase the power to identify specific signals in complex genomic datasets (they found "greatly sharpened global and gene-specific differential expression across treatment groups")

- But caution should be exercised to avoid removing biological signal of interest

- We must be precise and deliberate in the design and analysis of experiments and the resulting data, and also mindful of the limitations we impose with our own perspective

- Open data exploration is not possible after such supervised “cleaning”, because effects beyond those stipulated by the researcher may have been removed
<br>

#### Comparing data with and without surrogate variable correction

```{r correct_svs_in_datExpr, warning=FALSE}
# Taken from https://www.biostars.org/p/121489/#121500
correctDatExpr = function(datExpr, mod, svs) {
  X = cbind(mod, svs)
  Hat = solve(t(X) %*% X) %*% t(X)
  beta = (Hat %*% t(datExpr))
  rm(Hat)
  gc()
  P = ncol(mod)
  return(datExpr - t(as.matrix(X[,-c(1:P)]) %*% beta[-c(1:P),]))
}

pca_samples_before = datExpr %>% t %>% prcomp
pca_genes_before = datExpr %>% prcomp

# Correct
mod = model.matrix(~ Diagnosis, colData(dds))
svs = datMeta %>% dplyr::select(SV1:SV13) %>% as.matrix
datExpr_corrected = correctDatExpr(as.matrix(datExpr), mod, svs)

pca_samples_after = datExpr_corrected %>% t %>% prcomp
pca_genes_after = datExpr_corrected %>% prcomp

rm(correctDatExpr)
```

#### Samples
<br>

Removing batch effects has a big impact in the distribution of the samples, separating them by diagnosis relatively well just using the first principal component (although the separation is not nearly as good as with the Gandal dataset)
```{r pca_samples, warning=FALSE}
pca_samples_df = rbind(data.frame('ID'=colnames(datExpr), 'PC1'=pca_samples_before$x[,1],
                                  'PC2'=pca_samples_before$x[,2], 'corrected'=0),
                       data.frame('ID'=colnames(datExpr), 'PC1'=pca_samples_after$x[,1],
                                  'PC2'=pca_samples_after$x[,2], 'corrected'=1)) %>%
                 left_join(datMeta %>% mutate('ID'=rownames(datMeta)), by='ID')

ggplotly(pca_samples_df %>% ggplot(aes(PC1, PC2, color=Diagnosis)) + geom_point(aes(frame=corrected, id=ID), alpha=0.75) + 
         xlab(paste0('PC1 (corr=', round(cor(pca_samples_before$x[,1],pca_samples_after$x[,1]),2),
                     '). % Var explained: ', round(100*summary(pca_samples_before)$importance[2,1],1),' to ',
                     round(100*summary(pca_samples_after)$importance[2,1],1))) +
         ylab(paste0('PC2 (corr=', round(cor(pca_samples_before$x[,2],pca_samples_after$x[,2]),2),
                     '). % Var explained: ', round(100*summary(pca_samples_before)$importance[2,2],1),' to ',
                     round(100*summary(pca_samples_after)$importance[2,2],1))) +
         ggtitle('Samples') + theme_minimal())

rm(pca_samples_df)
```
<br>

#### Genes
<br>

It seems like the sva correction preserves the mean expression of the genes and erases almost everything else (although what little else remains is enough to characterise the two Diagnosis groups pretty well using only the first PC)

*Plot is done with only 10% of the genes so it's not that heavy
```{r pca_genes, warning=FALSE, message=FALSE}
pca_genes_df = rbind(data.frame('ID'=rownames(datExpr), 'PC1'=pca_genes_before$x[,1],
                                'PC2'=pca_genes_before$x[,2], 'corrected'=0, 'MeanExpr'=rowMeans(datExpr)),
                     data.frame('ID'=rownames(datExpr), 'PC1'=pca_genes_after$x[,1],
                                'PC2'=pca_genes_after$x[,2], 'corrected'=1, 'MeanExpr'=rowMeans(datExpr)))

keep_genes = rownames(datExpr) %>% sample(0.1*nrow(datExpr))

pca_genes_df = pca_genes_df %>% filter(ID %in% keep_genes)

ggplotly(pca_genes_df %>% ggplot(aes(PC1, PC2,color=MeanExpr)) + 
         geom_point(alpha=0.3, aes(frame=corrected, id=ID)) +
         xlab(paste0('PC1 (corr=', round(cor(pca_genes_before$x[,1],pca_genes_after$x[,1]),2),
                     '). % Var explained: ', round(100*summary(pca_genes_before)$importance[2,1],1),' to ',
                     round(100*summary(pca_genes_after)$importance[2,1],1))) +
         ylab(paste0('PC2 (corr=', round(cor(pca_genes_before$x[,2],pca_genes_after$x[,2]),2),
                     '). % Var explained: ', round(100*summary(pca_genes_before)$importance[2,2],1),' to ',
                     round(100*summary(pca_genes_after)$importance[2,2],1))) +
         scale_color_viridis() + ggtitle('Genes') + theme_minimal())


rm(pca_samples_before, pca_genes_before, mod, svs, pca_samples_after, pca_genes_after, pca_genes_df, keep_genes)
```


Everything looks good, so we're keeping the corrected expression dataset
```{r}
datExpr = datExpr_corrected

rm(datExpr_corrected)
```

<br>

### Processing Group
<br>

Even after correcting the dataset for the surrogate variables found with sva, there is still a difference in mean expression by processing, but this time, the Harvard Group has a higher level of expression than the Maryland one, which agrees with their ASD-Control sample imbalance

```{r, warning=FALSE, message=FALSE}
plot_data_b1 = data.frame('Mean'=colMeans(datExpr[,datMeta$SiteHM=='H']), 'Batch'='H')
plot_data_b2 = data.frame('Mean'=colMeans(datExpr[,datMeta$SiteHM=='M']), 'Batch'='M')

plot_data = rbind(plot_data_b1, plot_data_b2)
mu = plot_data %>% group_by(Batch) %>% dplyr::summarise(BatchMean=mean(Mean)) %>% ungroup

ggplotly(plot_data %>% ggplot(aes(x=Mean, color=Batch, fill=Batch)) + geom_density(alpha=0.3) + 
         geom_vline(data=mu, aes(xintercept=BatchMean, color=Batch), linetype='dashed') +
         ggtitle('Mean expression by sample grouped by processing date') + scale_x_log10() + theme_minimal())

rm(plot_data_b1, plot_data_b2, plot_data, mu)
```
<br>

### Performing Batch Correction for Processing Group

<br><br>

I will save the batch corrected dataset as a different dataset because of the correlation between processing group and diagnosis and I will use the dataset without the batch correction for the rest of the pipeline to avoid losing Diagnosis information when removing the Processing Group signal from the data

https://support.bioconductor.org/p/50983/

```{r ComBat}
datExpr_combat = datExpr %>% as.matrix %>% ComBat(batch=datMeta$SiteHM)
```

Now both batches have almost the same mean expression
```{r, warning=FALSE}
plot_data_b1 = data.frame('Mean'=colMeans(datExpr_combat[,datMeta$SiteHM=='H']), 'Batch'='H')
plot_data_b2 = data.frame('Mean'=colMeans(datExpr_combat[,datMeta$SiteHM=='M']), 'Batch'='M')

plot_data = rbind(plot_data_b1, plot_data_b2)
mu = plot_data %>% group_by(Batch) %>% dplyr::summarise(BatchMean=mean(Mean))

ggplotly(plot_data %>% ggplot(aes(x=Mean, color=Batch, fill=Batch)) + geom_density(alpha=0.3) + 
         geom_vline(data=mu, aes(xintercept=BatchMean, color=Batch), linetype='dashed') +
         ggtitle('Mean expression by sample grouped by processing date') + scale_x_log10() + theme_minimal())

rm(plot_data_b1, plot_data_b2, plot_data, mu)
```
<br>

---

#### Save preprocessed datasets with and without ComBat correction
```{r save_preprocessed_dataset}
save(datExpr, datMeta, datGenes, DE_info, dds, file='./../Data/preprocessed_data.RData')
save(datExpr_combat, datMeta, datGenes, DE_info, dds, file='./../Data/preprocessed_data_ComBat.RData')
```
<br><br>

---

#### Session info
```{r print_session_info}
sessionInfo()
```
<br><br>
