---
title: 'Perform WGCNA'
output:
  html_document:
    code_folding: 'hide'
---

```{r load_packages, echo=TRUE, warning=FALSE, message=FALSE}
#setwd('/afs/inf.ed.ac.uk/user/s17/s1725186/Documents/PhD-Models/FirstPUModel/RMarkdowns')

library(tidyverse) ; library(reshape2) ; library(glue) ; library(dendextend)
library(RColorBrewer) ; require(gridExtra) ; library(GGally)
suppressMessages(library(WGCNA))
library(ClusterR)
library(expss)
```

Load preprocessed dataset (preprocessing code in 01_data_preprocessing.Rmd)
```{r load_dataset, echo=TRUE, message=FALSE, warning=FALSE}

# Gandal dataset
load('./../Data/preprocessed_data.RData')
datExpr = datExpr %>% data.frame
DE_info = DE_info %>% data.frame

# GO Neuronal annotations: regex 'neuron' in GO functional annotations and label the genes that make a match as neuronal
GO_annotations = read.csv('./../Data/genes_GO_annotations.csv')
GO_neuronal = GO_annotations %>% filter(grepl('neuron', go_term)) %>% 
              mutate('ID'=as.character(ensembl_gene_id)) %>% 
              dplyr::select(-ensembl_gene_id) %>% distinct(ID) %>%
              mutate('Neuronal'=1)


# SFARI Genes
SFARI_genes = read_csv('./../../../SFARI/Data/SFARI_genes_01-03-2020_w_ensembl_IDs.csv')
SFARI_genes = SFARI_genes[!duplicated(SFARI_genes$ID) & !is.na(SFARI_genes$ID),]


# Update DE_info with SFARI and Neuronal information
DE_info = DE_info %>% mutate('ID'=rownames(.)) %>% left_join(SFARI_genes, by='ID') %>% 
  mutate(`gene-score`=ifelse(is.na(`gene-score`), 'None', `gene-score`)) %>%
  distinct(ID, .keep_all = TRUE) %>% left_join(GO_neuronal, by='ID') %>%
  mutate(Neuronal=ifelse(is.na(Neuronal), 0, Neuronal)) %>%
  mutate(gene.score=ifelse(`gene-score`=='None' & Neuronal==1, 'Neuronal', `gene-score`), significant=padj<0.05 & !is.na(padj))


# Get colors from the ggplot palette
gg_colour_hue = function(n) {
  hues = seq(15, 375, length = n+1)
  pal = hcl(h = hues, l = 65, c = 100)[1:n]
}

rm(GO_annotations)
```
<br>

# Perform Clustering
<br>

Pipeline summary:

- Using biweight correlation as correlation metric

- Using signed hybrid function to create the adjacency matrix

- Elevating the adjacency matrix to the best power to obtain a scale-free topology

- Using a Topological Overlap Matrix (TOM) as distance matrix

- Performing hierarchical clustering (using average linkage hclust(method='average'))

- Extracting clusters using the Dynamic Tree brach cutting algorithm from [Dynamic Tree Cut: in-depth description, tests and applications](https://horvath.genetics.ucla.edu/html/CoexpressionNetwork/BranchCutting/Supplement.pdf)

- Merging similar clusters using Module Eigengenes

- Saving results with different cluster merges so we can decide which to use later

<br><br>

---

### Define a gene co-expression similarity
<br>

Using Biweight midcorrelation because it's more robust to outliers than regular correlation or Mutual Information score

Correcting the correlation matrix from $s \in [-1,1]$ to $s \in [0,1]$. Three methods are proposed: 

1. Unsigned: $s_{ij}=|bw(i,j)|$

2. Signed: $s_{ij}=\frac{1+bw(i,j)}{2}$

3. Signed Hybrid: $s_{ij}=\frac{1+bw(i,j)}{2}$ if $cor(i,j)>0$ and $s_{ij}=0$ else

Notes on each method:

  - Using the unsigned adjacency we lose the direction of the correlation, but at least we maintain the magnitude of the correlation of all the genes

  - Using the signed adjacency, the strongest negative correlations (-1) get mapped to 0 (no correlation) and the zero correlated genes get mapped to the average correlation (0.5), which I don't think makes much sense
  
  - Using the signed hybrid adjacency, we lose all the negative correlations but we mantain the magnitude of the positive ones. This seems to be the best option since genes with negative correlations usually don't share the same biological pathways, so losing this information is not as important as maintaining the correct information for the genes with the positive correlation (and combining together these two types of correlation as the unsigned approach does seems to mix different behaviours together)
  
**Note:** According to [this post](https://peterlangfelder.com/2018/11/25/__trashed/) in practice there's not a big difference between signed and signed hybrid adjacencies because you end up elevating the values to a certain power, which in both cases moves the negative correlations to nearly zero.
  

<br><br>

---

### Define a family of adjacency functions
<br>

- Sigmoid function: $a(i,j)=sigmoid(s_{ij}, \alpha, \tau_0) \equiv \frac{1}{1+e^{-\alpha(s_{ij}-\tau_0)}}$

- Power adjacency function: $a(i,j)=power(s_{ij}, \beta) \equiv |S_{ij}|^\beta$

Chose power adjacency function over the sigmoid function because it has only one parameter to adjust and both methods are supposed to lead to very similar results if the parameters are chosen with the scale-free topology criterion.
<br>

### Choosing a parameter value
<br>

Following the **scale-free topology criterion** because metabolic networks have been found to display approximate scale free topology

1. Only consider those parameter values that lead to a network satisfying scale-free topology at least approximately, e.g. signed $R^2 > 0.80$

**Note:** Including non-integer powers in the parameters. Integer powers are usually used for this but there's no reason why you can't use a non-integer power and you can get a lower power vector that is above the 0.8 threshold so you don't affect the data that much

```{r best_power}
allowWGCNAThreads()

best_power = datExpr %>% t %>% pickSoftThreshold(powerVector = c(1,2,3,4,5, seq(5.5,6,0.1)), RsquaredCut=0.8)
```
Best power for scale free topology: `r  best_power$powerEstimate`
<br>

```{r create_adj_mat}
S_sft = datExpr %>% t %>% adjacency(type='signed hybrid', power=best_power$powerEstimate, corFnc='bicor')
```
<br><br>

---

## Defining a measure of node dissimilarity
<br>

Using topological overlap dissimilarity measure because it has been found to result in biologically meaningful modules

```{r TOM}
TOM = S_sft %>% TOMsimilarity
rownames(TOM) = rownames(S_sft)
colnames(TOM) = colnames(S_sft)

#library(igraph)
#g = graph_from_adjacency_matrix(TOM, weighted=TRUE, mode='undirected')
#edges = as_edgelist(g)
#e = cbind(edges, E(g)$weight)

dissTOM = 1-TOM

rm(S_sft, TOM)
```

<br><br>

---

## Identifying gene modules

Using hierarchical clustering using average linkage on the TOM-based dissimilarity matrix

```{r, fig.width=10}
dend = dissTOM %>% as.dist %>% hclust(method='average')
plot(dend, hang=0, labels=FALSE)
```


Instead of using a fixed height to cut the dendrogram into clusters, using a dynamic branch cutting approach taken from [Dynamic Tree Cut: in-depth description, tests and applications](https://horvath.genetics.ucla.edu/html/CoexpressionNetwork/BranchCutting/Supplement.pdf)

There are two algorithms that do this:

**Dynamic Tree Cut:** top-down algorithm relying only on the dendrogram and respecting the order of the clustered objects on it. This method is less sensitive to parameter choice but also less flexible and it performed better in our previous experiments **but** when using it on this dataset it left most genes without a cluster, so tried doing it also with the Dynamic Hybrid algorithm

 **Dynamic Hybrid Cut:** builds the clusters from bottom up. In addition to information from the dendrogram, it utilizes dissimilarity information among the objects. Seems to me that relies on too many heuristics and has too many parameters to tune. Ran it with the default settings
<br>

### Dynamic Tree Cut Algorithm
<br>

This algorithm leaves a lot of genes without a cluster, specially genes close to the root of the dendrogram which is uncertain how to assign. I'll save the results for this algorithm for completeness, but will use the output from Dynamic Hybrid Cut for the rest of the analysis

```{r}
modules = cutreeDynamic(dend, method = 'tree', minClusterSize = 10)
```

`r sum(modules==0)` genes (`r round(100*mean(modules==0))`%)  are left without a cluster

```{r}
table(modules)
# Note: The modules are ordered as in the rows in datExpr
```
<br>

#### Merging similar modules
<br>

```{r merge_modules_dynamic_tree, fig.width=10}

# Calculate eigengenes
MEList = datExpr %>% t %>% moduleEigengenes(colors = modules)
MEs = MEList$eigengenes

# Calculate dissimilarity of module eigengenes
MEDiss = 1-cor(MEs)

# Cluster module eigengenes
METree = hclust(as.dist(MEDiss), method = 'average')

METree %>% as.dendrogram %>% plot(main = 'Clustering of module eigengenes', leaflab = 'none')
abline(h=1, col='#0099cc')
abline(h=0.5, col='#009999')
abline(h=0.3, col='#99cc00')

merge_top = datExpr %>% t %>% mergeCloseModules(modules, cutHeight = 1)

merge_similar = datExpr %>% t %>% mergeCloseModules(modules, cutHeight = 0.5)

merge_most_similar = datExpr %>% t %>% mergeCloseModules(modules, cutHeight = 0.3)

rm(MEList, MEs, MEDiss, METree)
```

#### Module assignment
<br>

Almost all of the genes except the ones farthest away from the root remain unclassified

```{r plot_dend_dynamic_tree, warning=FALSE, fig.width=10}

get_mod_colours = function(mods){
  
  n = length(unique(mods))-1
  set.seed(123) ; rand_order = sample(1:n)
  mod_colors = c('gray',gg_colour_hue(n)[rand_order])
  names(mod_colors) = mods %>% table %>% names
  
  return(mod_colors)
}

module_colors = get_mod_colours(modules)
merged_sim_module_colors = get_mod_colours(merge_most_similar$colors)
merged_module_colors = get_mod_colours(merge_similar$colors)
top_module_colors = get_mod_colours(merge_top$colors)

dend_colors = data.frame('ID' = rownames(datExpr),
                         'OriginalModules' = module_colors[as.character(modules)],
                         'MostSimModules' = merged_sim_module_colors[as.character(merge_most_similar$colors)],
                         'MergedModules' = merged_module_colors[as.character(merge_similar$colors)],
                         'TopModules' = top_module_colors[as.character(merge_top$colors)])

dend %>% as.dendrogram(hang=0) %>% plot(ylim=c(min(dend$height),1), leaflab='none')
colored_bars(colors=dend_colors[dend$order,-1])

rm(module_colors, merged_sim_module_colors, merged_module_colors, top_module_colors)
```

```{r}
modules_dynamic_tree = dend_colors

rm(dend_colors)
```
<br>

### Dynamic Hybrid Cut Algorithm

```{r}
modules = cutreeDynamic(dend, minClusterSize = 10, distM = dissTOM)
```

`r sum(modules==0)` genes (`r round(100*mean(modules==0),2)`%)  are left without a cluster

```{r}
table(modules)
```
<br>

#### Merging similar modules
<br>

```{r merge_modules_dynamic_hybrid, fig.width=10}

# Calculate eigengenes
MEList = datExpr %>% t %>% moduleEigengenes(colors = modules)
MEs = MEList$eigengenes

# Calculate dissimilarity of module eigengenes
MEDiss = 1-cor(MEs)

# Cluster module eigengenes
METree = hclust(as.dist(MEDiss), method = 'average')

METree %>% as.dendrogram %>% plot(main = 'Clustering of module eigengenes', leaflab = 'none')
abline(h=1, col='#0099cc')
abline(h=0.5, col='#009999')
abline(h=0.3, col='#99cc00')

merge_top = datExpr %>% t %>% mergeCloseModules(modules, cutHeight = 1)

merge_similar = datExpr %>% t %>% mergeCloseModules(modules, cutHeight = 0.5)

merge_most_similar = datExpr %>% t %>% mergeCloseModules(modules, cutHeight = 0.3)

rm(MEList, MEs, MEDiss, METree)
```

#### Module assignment
<br>

Classification is noisy with respect to the dendrogram, but that isn't necessarily a bad thing, as Peter Langfelder explains in [this blog post ](https://peterlangfelder.com/2018/12/30/why-wgcna-modules-dont-always-agree-with-the-dendrogram/) from [Insights from a Million Correlations](https://peterlangfelder.com/articles/)

```{r plot_dend_dynamic_hybrid, warning=FALSE, fig.width=10}
module_colors = get_mod_colours(modules)
merged_sim_module_colors = get_mod_colours(merge_most_similar$colors)
merged_module_colors = get_mod_colours(merge_similar$colors)
top_module_colors = get_mod_colours(merge_top$colors)

dend_colors = data.frame('ID' = rownames(datExpr),
                         'OriginalModules' = module_colors[as.character(modules)],
                         'MostSimModules' = merged_sim_module_colors[as.character(merge_most_similar$colors)],
                         'MergedModules' = merged_module_colors[as.character(merge_similar$colors)],
                         'TopModules' = top_module_colors[as.character(merge_top$colors)])

dend %>% as.dendrogram(hang=0) %>% plot(ylim=c(min(dend$height),1), leaflab='none')
colored_bars(colors=dend_colors[dend$order,-1])

rm(module_colors, merged_module_colors, top_module_colors)
```

```{r}
modules_dynamic_hybrid = dend_colors

rm(dend_colors)
```

---

Save clusterings file
```{r}
colnames(modules_dynamic_tree) = c('ID','DynamicTree', 'DynamicTreeMergedSmall', 'DynamicTreeMerged', 'DynamicTreeTop')
colnames(modules_dynamic_hybrid) = c('ID','DynamicHybrid', 'DynamicHybridMergedSmall', 'DynamicHybridMerged', 'DynamicHybridTop')

modules = modules_dynamic_tree %>% left_join(modules_dynamic_hybrid, by='ID')

write.csv(modules, file='./../Data/clusters.csv', row.names=FALSE)
```
<br><br>

---

#### Session info
```{r print_session_info}
sessionInfo()
```
<br><br>
